---
title: "Heart_Disease_Analysis"
author: "Ocansey Djabaku Kevin"
date: '2022-08-12'
output: html_document
editor_options: 
  markdown: 
    wrap: 76
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
set.seed(101)

library(tidyverse)
library(dplyr)
library(dslabs)
library(caret)
library(purrr)
```

## Reading heart.csv

```{r heart}
heart <- read.csv(file.choose(), header=TRUE)

# age
# sex
# cp = chest pain type (4 values)
# trestbps = resting blood pressure measured in mm Hg on admission to the hospital.
# chol = serum cholestoral in mg/dl  
# fbs = fasting blood sugar > 120 mg/dl
# restecg = resting electrocardiographic results (values 0= normal,1 = having ST-T wave abnormality, 2 = showing probable or definite left ventricular hypertropy by Estes )  
# thalach = maximum heart rate achieved
# exang = exercise induced angina
# oldpeak = ST depression induced by exercise relative to rest
# slope = the slope of the peak exercise ST segment; 1 = upslopping 2 = flat 3 = downslopping 
# ca = number of major vessels (0-3) colored by flourosopy
# thal: 0 = normal; 1 = fixed defect; 2 = reversible defect
# target - the predicted attribute - diagnosis of heart disease ( angiographic disease status) value 0 <= 50% diameter narrowing; and value 1  => 50% diameter narrowing 
# The names and social security numbers of the patients were recently removed from the database, replaced with dummy values.
```

## CLEANING DATA

```{r Cleaning Data}

#  After close look at each attribute i found discrepancies for two predictors. Ca and Thal.
# Following the description provided from Kaggle, recurring values, ca-4 and thal-0 were not accounted for.
# So either the mistake was made dUring the description of the data set or made during the collection.


  heart$ca <- factor(heart$ca)
  heart$thal <-factor(heart$thal)


 heart_clean <- heart %>% filter(ca != 4) %>% filter(thal != 0)
 heart_clean <- heart_clean %>% mutate(ca = droplevels(ca), thal = droplevels(thal))
 
 head(heart_clean)
```

## REFIXING DATA VARIABLES FOR VISUALIZATION

```{r creating heart2}

# We are going to use this new data frame "heart 2" to do some visualization,
# hence we need to change the levels of the values into readable characters


heart2 <-  heart_clean %>% mutate(sex = if_else(sex == 1, "Male", "Female"),
                            age = if_else(age <= 64 & age >= 24, "Adult", "Elderly"),
                  fbs = if_else( fbs == 1, ">120", "<=120"),
                  exang = if_else(exang ==  1, "Yes", "No"),
                  cp =if_else(cp == 1, "Typical Angina Pain", 
                              if_else(cp == 2,  "Atypical Angina Pain", 
                                     ifelse( cp == 3, "Non-Anginal Pain", "Asymptomatic Pain"))),
                  restecg = ifelse(restecg == 0, "Normal", if_else(restecg ==  1, "Abnormality", "Probable or definite")),
                  slope = if_else( slope == 1,"upslopping", if_else(slope == 2, "Flat", "Downslopping")),
                  ca = as.factor(ca),
                  thal = if_else( thal == 1, "Fixed defect", if_else(thal == 2, "Normal", "Reversible Defect")),
                  target = if_else(target == 1,"Yes", "No")) %>% mutate_if(is.character, as.factor) %>% select(target, sex, fbs, exang, cp, restecg, slope, thal, everything())

head(heart2)
```

## EXPLORATORY DATA ANALYSIS

```{r EXPLORATORY DATA}

head(heart_clean)
summary(heart2)

```

## DATA VISUALIZATION

```{r heart2}
heart2 %>% ggplot(aes(target , fill = target)) + 
  geom_bar() + 
  xlab("Heart Disease") + 
  ggtitle("Presence Of Heart Disease") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Heart disease analysis with Sex attribute

```{r heart2}
# Heart disease Grouped by Sex 2
heart2 %>% filter(target == "Yes") %>%  group_by(sex) %>% count() %>%
  ggplot(aes(sex, n, fill = sex)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Sex") +
  xlab("Sex") + ylab("count") 


# Check to see if Males are prevalent in our data.
prop.table(table(heart2$sex))
# There are 2 times more Males than females in the dataset. something to keep in mind 
```

```{r heart2}
# Looking at the graph above we cannot make a general statement that Males are more likely to suffer from heart diseases,
# because from the data there is prevalence of Males
# Hence we can check the proportion of Heart disease for each gender
prop.table(table(heart2$sex, heart2$target))

# We can represent or findings in a graph like so
heart2 %>% group_by(sex, target) %>% count() %>%
  ggplot(aes(sex, n, fill = target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Sex") +
  xlab("Sex") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))

# We notice although there are more men with heart Disease in the data set generally. it seems Females have a higher concentration of heart disease.
```

### Sex and Age

```{r heart2}
heart2 %>% group_by(sex, age, target) %>% count() %>%
  ggplot(aes(sex, n, fill = target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Sex") +
  xlab("Sex") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present")) + facet_wrap(~age)

# This graph shows Heart Disease are prevalent in females across our various ages as well.
```

### Heart disease analysis with Chol

```{r Cholesterol}
# Comparing the correlation between cholesterol levels and heart disease 
heart2 %>% group_by(chol,target) %>% count() %>%  ggplot(aes(chol,n)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis by Chol") +
  xlab("Cholesterol") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present")) + facet_wrap(~target)

# Zooming on the cholesterol levels between 200 and 300.
heart2 %>% filter(chol > 200 & chol < 300) %>% group_by(chol,target) %>% count() %>%  ggplot(aes(chol,n)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Cholesterol") +
  xlab("Cholesterol") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present")) + facet_wrap(~target)

heart2 %>% filter(chol > 200 & chol < 300) %>% group_by(target) %>% summarize(n())
```

### Fast Blood Sugar

```{r Fast Blood Sugar}

heart2 %>% group_by(fbs,target) %>% count() %>%  ggplot(aes(fbs,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Fasting Blood Sugar") +
  xlab("FBS") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present")) 
```

### Chest Pain

```{r # Chest Pain}
heart2 %>% group_by(cp,target) %>% count() %>%  ggplot(aes(cp,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Chest Pain") +
  xlab("Chest Pain") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))    
```

### Ca

```{r Ca }

heart2 %>% group_by(ca,target) %>% count() %>%  ggplot(aes(ca,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with CA") +
  xlab("CA") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Thal

```{r Thal}
heart2 %>% group_by(thal,target) %>% count() %>%  ggplot(aes(thal,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Thal") +
  xlab("THAL") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Thalach

```{r Thalach}
heart2 %>% group_by(thalach,target) %>% count() %>%  ggplot(aes(thalach,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Thalach") +
  xlab("THALACH") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))

```

### Trestbps

```{r Trestbps}
heart2 %>% group_by(trestbps,target) %>% count() %>%  ggplot(aes(trestbps,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Trestbps") +
  xlab("Trestbps") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Slope

```{r Slope}
heart2 %>% group_by(slope,target) %>% count() %>%  ggplot(aes(slope,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Slope") +
  xlab("Slope") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Oldpeak

```{r Oldpeak }
heart2 %>% group_by(oldpeak,target) %>% count() %>%  ggplot(aes(oldpeak,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with OldPeak") +
  xlab("Oldpeak") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Exang

```{r Exang}
heart2 %>% group_by(exang,target) %>% count() %>%  ggplot(aes(exang,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Exang") +
  xlab("Exang") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

### Restecg

```{r Restecg}
heart2 %>% group_by(restecg,target) %>% count() %>%  ggplot(aes(restecg,n, fill= target)) +
  geom_col()  +
  ggtitle("Heart Disease Analysis with Restecg") +
  xlab("Restecg") + ylab("count") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absent", "Present"))
```

# MACHINE LEARNING

### Fix the variable types (making sure they are factors)

```{r Fix}
heart_clean$sex <-factor(heart_clean$sex)
heart_clean$cp<- factor(heart_clean$cp)
heart_clean$fbs <- factor(heart_clean$fbs)
heart_clean$restecg <- factor(heart_clean$restecg)
heart_clean$exang <- factor(heart_clean$exang)
heart_clean$slope <- factor(heart_clean$slope)
heart_clean$ca <- factor(heart_clean$ca)
heart_clean$thal <-factor(heart_clean$thal)
heart_clean$target <- factor(heart_clean$target)
```

### Split the Data

```{r Data Partitioning}
set.seed(101)
y <- heart_clean$target
test_index <- createDataPartition(y,times = 1, p = 0.8, list = FALSE)

train_set <- heart_clean[test_index,]
test_set <- heart_clean[-test_index,]
```

### Simple model based on sex

```{r Simple Model train}
prop.table(table(heart2$sex, heart2$target))  
# 86 females do not have heart disease. whiles 223 females out of a total 319 do. 
# that means, according to our dataset 73 out of 100 Females are likely to have heart disease. 
# Only 41 percent of the males have heart disease.

#  Considering this Lets try a simple model where we predict if a patient has heart disease based on their sex
# target is mutated to be 1 if  sex = 0  which is Female in our dataset.
# We chose Females because according to our proportion  males are less likely to have heart diseases.

y_hat_sex <- factor(ifelse(heart_clean$sex == 0, 1, 0))
confusionMatrix(y_hat_sex, heart_clean$target)$overall["Accuracy"]
# We get an accuracy of 0.629. Not bad, but we applied the model still using the main data set. 
```

```{r Simple Model test}
# Applying this Model on the Test set
y_hat_sex_test <- factor(ifelse(test_set$sex == 0, 1, 0))
confusionMatrix(y_hat_sex_test, test_set$target)$overall["Accuracy"]
# we get an accuracy of 0.587  
```

### USING GLM

```{r}
# considering conditional probabilities we want to determine the best p value for improving our accuracy
set.seed(101)
ps <- seq(0.3, 0.7, 0.1)

accuracy_general <- map_df(ps, function(p){
  
fit <- glm(target ~ exang + trestbps + chol + age + sex + thal + restecg + cp + fbs + ca + oldpeak + slope + thalach, data = train_set, family = binomial)
p_hat <- predict(fit, test_set, type = "response")
y_hat <-  factor(ifelse(p_hat > p, 1, 0))
cm <- confusionMatrix(y_hat, as.factor(test_set$target))
acc <- cm$overall["Accuracy"]

f_meas <- F_meas(y_hat, test_set$target)
cm_sensitivity <- sensitivity(y_hat, test_set$target)
cm_specificity <- specificity(y_hat, test_set$target)


tibble( acc, f_meas, cm_sensitivity, cm_specificity)
})  
 
accuracy_results <- cbind( ps, accuracy_general)
accuracy_results
# You can decide which p value cutoff to use. depending on F meas and cm sensitivity 0.6 seems to be the best.
```

### IMPROVING ON OUR PREVIOUS MODEL USING K-FOLD CROSS VALIDATION

```{r Glm and K-cross Validation}
set.seed(101)
  train_control  <- trainControl(method = 'cv', number = 10)
  
  #after a couple permutations the final attributes/predictors to be used for this model were tracked using varImp()
  
  model_glm <- train(target ~  thal +  cp  + exang + trestbps +  sex + chol + ca + thal + oldpeak +slope+ thalach, data = train_set, family = binomial, method = "glm", trControl = train_control)
  y_hat_glm <- predict(model_glm, test_set)
  cm_glm <-  confusionMatrix(y_hat_glm,test_set$target)
  cm_glm$overall["Accuracy"]
  F_meas(y_hat_glm,test_set$target)

  im_glm <- varImp(model_glm)
  im_glm
  
  #  We get an accuracy of 0.8542714 and an F_meas of  0.8465608
```

### USING KNN

```{r knn}

  set.seed(101)
  #after a couple permutations the final attributes/predictors to be used for this model were tracked using varImp(model_knn)
  
  model_knn <- train(target ~  oldpeak + thalach + ca + thal , data = train_set,  method = "knn", trControl = train_control)
  y_hat_knn <- predict(model_knn, test_set)
  cm_knn <-  confusionMatrix(y_hat_knn,test_set$target)
  cm_knn$overall["Accuracy"]

  im_knn <- varImp(model_knn)
  im_knn  
  # i noticed thal has a varImp of 0, however if we remove the predictor the total accuracy drops.




```

### CLASSIFICATION TREEE

```{r Classification Tree}
  library(rpart)
  set.seed(101)
  
  #after a couple permutations the final attributes/predictors to be used for this model were tracked using     varImp(tree_model)
  tree_model <- rpart(target ~  exang + thal +  cp  + ca + oldpeak +  thalach, data = train_set, method = "class")
  varImp(tree_model)
  
  plotcp(tree_model)
  plot(tree_model, uniform=TRUE,
       main="Classification Tree for Heart Disease")
  text(tree_model, use.n = TRUE, all=TRUE, cex=.75)
  
  y_hat_tree <- predict(tree_model, test_set, type = 'class')
  cm_tree <- confusionMatrix(y_hat_tree, test_set$target)
```

### RANDOM FOREST

```{r Random Forest}
  library(randomForest)
  set.seed(101)
  model_forest <- randomForest(target ~ exang + trestbps + chol + age + sex + thal + restecg + cp + fbs + 
                           ca + oldpeak + slope + thalach, data = train_set, ntree = 50)
  y_hat_forest <- predict(model_forest, test_set, type = 'class')
  cm_forest <- confusionMatrix(y_hat_forest, test_set$target)
  cm_forest
  varImp(model_forest)
  
  # We get a high accuracy between 0.98 and 1
```
